Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-02-06 papers/LOPe/transformations.tex
Beta rough draft.

\section{LOPe Implementation}

@@

@@

Amortization of halo transfer and kernel

>>>>

\section{LOPe and CAFe Implementation}

@@

\subsection{Source-to-Source Transformation Infrastructure}

1. Parsing to produce an AST nodes representated as ATerms.
2. Transformations on the AST nodes represented as rewrite rules and strategies in Stratego/XT
3. Pretty printing of transformed nodes using the pretty printing facilites of BoxLib.

\subsection{Transformations for Concurrent Procedures}

A key component of the LOPe transformations is the targeting of a LOPe \texttt{CONCURRENT}
procedure for a particular hardware (and software) architecture.  The execution target can be one
of several choices, including serial execution by the current process (and hardware) via inlining
of the function, parallel execution by inlining with OpenMP compiler directives, or parallel
execution by a heterogeneous processing elements with a language like OpenCL, for example.

For this work we have developed rewrite strategies and rules in Stratego/XT to rewrite Fortran AST
nodes to C AST nodes (extended with necessary OpenCL keywords).  The C AST ATerms have mostly a
one-to-one correspondence with Fortran terms.  A \texttt{CONCURRENT} procedure is transformed to a
OpenCL kernel (with the \texttt{\_\_kernel} attribute), Fortran formal parameters are transformed
to C parameters (with a direct mapping of types), and local Fortran variable declarations are rewritten
as C declarations.  Similarly, Fortran executable statements are rewritten as C executable AST nodes.
Mostly the only complication is mapping the LOPe local array index view to the global C index space.
This translation is facilitated by the Fortran symbol table that stores array shape and halo
size information.

Once all rewriting passes have been made to transform from LOPe Fortran ATerms to C ATerms, the
C ATerms are output as an OpenCL file using a C pretty printer.

\subsection{Transformations at the Call Site}

Transformations at a LOPe procedure call site are more difficult, though technically straight forward.
The Fortran function call must be transformed to a call to run the OpenCL kernel (produced as described
above).  This is facilitated by the use of the ForOpenCL library that provides Fortran bindings to the
OpenCL runtime\cite{foropencl}.  But the usage requires the declaration of extra variables,
allocation of memory on the OpenCL device (subimage), transfer of memory, marshalling of kernel
arguments, and synchronization.

These transformations are accomplished using several rewrite stages using Stratego/XT strategies:
1. A symbol table is produced in the first pass to store information related to arrays including,
array shape, halo size, and allocation status;
2. Additional variables are declared that are used to maintain the OpenCL runtime state, including
the OpenCL device (as obtained from a \texttt{GET\_SUBIMAGE} call), the OpenCL kernel (produced as
described above), and OpenCL variables used to marshall kernel arguments; and
3. All CAFe code related to subimage usage is lowered to standard Fortran with calls to the
ForOpenCL library as needed.  The latter step includes the allocation of OpenCL device memory,
transfer of memory to and from the OpenCL device and the executing Fortran process (image) as
indicated in the program source by the single bracket \texttt{[subimage]} notation, marshalling
of kernel argument, running of the OpenCL kernel, and synchronization.

Though not yet available, similar rewrite strategies are planned for targeting programming models
other than OpenCL including parallel execution with OpenMP directives.  In addition, simple serial
execution with function inlining (if desired) will be performed by the regular Fortran compiler once
all CAFe and LOPe code has been lowered to standard Fortran.

@@

A thorough examination of potential performance gains using LOPe for parallelization of code is beyond
the scope of this paper.  The emphasis of this work is to explain the syntax and execution symantics
of a LOPe application.  However, preliminary work using the StrategoXT rewrite system as described indicates
that similar performance can be expected to that of previous work\cite{foropencl} (see also\cite{StencilCompilers}).

The current implementation does \emph{not} take advantage of optimization strategies such as
prefetching of array tiles (including halos) into OpenCL local memory or of loop unrolling on the
OpenCL device.  Neither does it take advantage of the potential of LOPe \texttt{CONCURRENT} function
fusion along with execution and synchronization strategies to overlap communication with
computation.

Since many scientific codes are dominated by memory performance, including and especially stencil algorithms
as they typically only involve a computation on a small locally central array element and a small overlapping
halo region.  Stencil operations frequently do not contain enough floating point operations per memory load
to allow for floating point performation to operate at peak (though this is entirely application and domain
specific).  Thus we illustrate the \emph{potential} for performance by noting the latency and throughput
performance of an attached GPU inconjuction with MPI distributed memory performance associated with halo
transfer in Table 1.

The results in Table 1 indicate that the primary bottleneck in using accelerators attached to the
TODO bus using OpenCL is likely to be the latency in transferring memory to and from the device for
distributed memory clusters of only a few nodes exchanging halo data using MPI.
